---
title: "Pre-requisites for Stable Isotope Mixing Models course"
author: "Andrew Parnell and Andrew Jackson"
output: html_document
---

In preparation for the course please install the following, preferably in the below suggested order. Make sure you run these before you arrive in case of poor (or crowded!) internet.

Remember you will need to bring your own personal computer for the duration of the course.

### Step 1

Install the following using the corresponding links and *make sure you have the latest versions installed*:

-	R: [http://www.r-project.org](http://www.r-project.org)

-	Rstudio (optional but recommended): [https://www.rstudio.com](https://www.rstudio.com)

-	JAGS: http://sourceforge.net/projects/mcmc-jags/files/

### Step 2

Now install all the packages required for the course by entering the following code in to R

```{r,eval=FALSE}
install.packages(c('rjags', 'compositions', 'devtools', 'tidyverse', 'vegan', 'ellipse'))

library(rjags)
library(compositions)
library(devtools) 
library(tidyverse)

install_github("andrewljackson/siar") 
library(siar)

install_github("andrewcparnell/simmr", build_vignettes = TRUE) 
library(simmr)

install_github("andrewljackson/SIBER", build_vignettes = TRUE) 
library(SIBER)

install_github("TGuillerme/mulTree@release")
install_github("healyke/SIDER", build_vignettes = TRUE) 
library(SIDER)

install.packages('MixSIAR')


```

These will install the most up to date versions of the packages. Note that this might take a while as some of these are pretty big and complicated packages.

 
### Step 3
 
Finally if you would like to do some pre-course reading (recommended) there are some papers which can be accessed [here](https://github.com/andrewcparnell/simms_course/tree/master/papers).


### Troubleshooting

Here are some common problems with solutions:

  - If you are on Windows and you get an error about not having Rtools please install Rtools from the link it provides you in the error message
  - If you are trying to install Rtools with R 3.5.0 and are getting an error message about incompatibility of Rtools, run the command:
```{r}
library(devtools)
assignInNamespace("version_info", c(devtools:::version_info, list("3.5" = list(version_min = "3.3.0", version_max = "99.99.99", path = "bin"))), "devtools")
```
  Then try the `install_github` command again.
  
  - If you are on a Mac you might need to install X11 from [http://www.xquartz.org](http://www.xquartz.org). 
  - You may get a pop up window be asking if you want 'additional build tools'. Please select 'yes'. 
  - If on a Mac you will likely need to install Xcode from the Mac App Store. This is a big download so make sure you download it before you get to the course.
  - If you get an error about package `stringi` try typing `install.packages('stringi',type='win.binary')` if on Windows or `install.packages('stringi',type='mac.binary')` if on Mac. Then re-run the commands above
  - If you get some errors about `pandoc` when installing simmr leave out the `build_vignettes=TRUE` argument


If you run into any other problems please drop us a line at <andrew.parnell@ucd.ie> (for SIAR/simmr/MixSIAR queries) or <a.jackson@tcd.ie> (for SIBER/SIDER queries).

