---
title: "Pre-requisites for Stable Isotope Mixing Models course"
author: "Andrew Parnell and Andrew Jackson"
output: html_document
---

In preparation for the course please install the following, preferably in the below suggested order. Make sure you run these before you arrive in case of poor (or crowded!) internet.

Remember you will need to bring your own personal computer for the duration of the course.

### Step 1

Install the following using the corresponding links and *make sure you have the latest versions installed*:

-	R: [http://www.r-project.org](http://www.r-project.org)

-	Rstudio (optional but recommended): [https://www.rstudio.com](https://www.rstudio.com)

-	JAGS: http://sourceforge.net/projects/mcmc-jags/files/ **_This step is really important and easy to skip over. JAGS is required, and is not installed as part of the R package rjags which is just a set of functions for R to interact with JAGS._**

### Step 2

Now install all the packages required for the course by entering the following code in to R

```{r,eval=FALSE}
install.packages(c('rjags', 'compositions', 'devtools', 'tidyverse', 'vegan', 'ellipse'))

# check that they load
library(rjags)
library(compositions)
library(devtools) 
library(tidyverse)

# Install the three main SIA packages
install.packages(c("simmr", "SIBER", "MixSIAR"))

# check that they load
library(simmr)
library(SIBER)
library(MixSIAR)

# SIDER is slightly different as it is not hosted on CRAN and 
# instead needs to be installed from GitHub
install_github("TGuillerme/mulTree@release")
install_github("healyke/SIDER", build_vignettes = TRUE) 

# check SIDER loads
library(SIDER)

# NB SIAR is out of date and we will not use it, but some of the example
# datasets are still in it and we need access to them.
install_github("andrewljackson/siar@master", build_vingettes = TRUE)




```

These will install the most up to date versions of the packages. Note that this might take a while as some of these are pretty big and complicated packages.

 
### Step 3
 
Finally if you would like to do some pre-course reading (recommended) there are some papers which can be accessed [here](https://github.com/andrewcparnell/simms_course/tree/master/papers).


### Troubleshooting

Here are some common problems with solutions:

#### Windows issues

  - If you are on Windows and you get an error about not having Rtools please install Rtools from the link it provides you in the error message
  - If you are trying to install Rtools with R 3.5.0 and are getting an error message about incompatibility of Rtools, run the command:
```{r, eval=FALSE}
library(devtools)
assignInNamespace("version_info", c(devtools:::version_info, list("3.5" = list(version_min = "3.3.0", version_max = "99.99.99", path = "bin"))), "devtools")
```
Then try the `install_github` command again.
- If you get an error about package `stringi` try typing `install.packages('stringi',type='win.binary')` then re-run the commands above.
  

#### Mac issues

  - you will more than likely need to install the command line developer tools. Instructions for this are available from (http://osxdaily.com/2014/02/12/install-command-line-tools-mac-os-x/). There is a larger set of tools called Xcode which is available from the Apple Store, but this is unnecessarily large and we dont recommend it unless you know you need it.
  - related to the installation of the command line tools, you may get errors such as `make: gfortran: no such file or directory mac` which relate to missing information for the calling of C++ and fortran compilers. This is likely remedied by running `sudo installer -pkg /Library/Developer/CommandLineTools/Packages/macOS_SDK_headers_for_macOS_10.14.pkg -target /` from a terminal window in OS X (not the R command window). This issue and solution is detailed in this thread: (https://github.com/r-lib/devtools/issues/1879#issuecomment-426426380)
  - If you are on a Mac you might need to install X11 from [http://www.xquartz.org](http://www.xquartz.org). 
  - If you get an error about package `stringi` try typing `install.packages('stringi',type='mac.binary')`. Then re-run the commands above
  <!-- - If you get some errors about `pandoc` when installing simmr leave out the `build_vignettes=TRUE` argument -->


If you run into any other problems please drop us a line at <andrew.parnell@mu.ie> (for SIAR/simmr/MixSIAR queries) or <a.jackson@tcd.ie> (for SIBER/SIDER queries).

