---
title: "Pre-requisites for Stable Isotope Mixing Models course"
author: "Andrew Parnell and Andrew Jackson"
output: html_document
---

In preparation for the course please install the following, preferably in the below suggested order. Make sure you run these as soon as possible to avoid falling behind.

### Step 1

Install the following using the corresponding links and *make sure you have the latest versions installed*:

-	R: [http://www.r-project.org](http://www.r-project.org): current version is v4.2.2 and we will assume this version throughout this course. If you want to run an older version or a custom setup then we may not be able to help with compatability issues that may arise.

-	Rstudio (optional but recommended): [https://www.rstudio.com](https://www.rstudio.com) - current version 2022.12.0+353

-	JAGS: http://sourceforge.net/projects/mcmc-jags/files/ **_This step is really important and easy to skip over. JAGS is required, and is not installed as part of the R package rjags which is just a set of functions for R to interact with JAGS._**

- Rtools for windows. If you are a windows user, it is likely you will need to install this suite of developer tools to install some custom packages https://cran.r-project.org/bin/windows/Rtools/rtools42/rtools.html

### Step 2

Now install all the packages required for the course by entering the following code in to R

```{r,eval=FALSE}
install.packages(c('rjags', 'R2jags','compositions', 'devtools', 
                   'tidyverse', 'vegan', 'ellipse'))

# check that they load
library(rjags)
library(R2jags)
library(compositions)
library(devtools) 
library(tidyverse)

# Install the three main SIA packages
install.packages(c("simmr", "SIBER", "MixSIAR"))

# check that they load
library(simmr)
library(SIBER)
library(MixSIAR)

# SIDER is slightly different as it is not hosted on CRAN and 
# instead needs to be installed from GitHub
install_github("TGuillerme/mulTree@release")
install_github("healyke/SIDER", build_vignettes = TRUE) 

# check SIDER loads
library(SIDER)

```

These will install the most up to date versions of the packages. Note that this might take a while as some of these are pretty big and complicated packages.
 
### Step 3
 
Finally if you would like to do some pre-course reading (recommended) there are some papers which can be accessed [here](https://github.com/andrewcparnell/simms_course/tree/master/papers).

Andrew Jackson has a youtube channel with [podcasts](https://www.youtube.com/playlist?list=PLXXzyqr0tEhPRNDuEQ6wcvMcpAkGRjVqs) on introduction to using R for linear modelling. There are links in the description to an associated repository of scripts and examples to go with it.  

Here is a [simple script for getting started with ggplot](https://andrewcparnell.github.io/simms_course/ap_notes/gg_plot_class.R) which might be useful for some but is not compulsory. 


### Troubleshooting

Here are some common problems with solutions:

#### Windows issues

  - If you are on Windows and you get an error about not having Rtools please install Rtools from the link it provides you in the error message or from here https://cran.r-project.org/bin/windows/Rtools/rtools42/rtools.html
  - If you are trying to install Rtools with R 3.5.0 (but you _really_ should be using >4.2) and are getting an error message about incompatibility of Rtools, run the command:
```{r, eval=FALSE}
library(devtools)
assignInNamespace("version_info", 
                  c(devtools:::version_info, 
                    list("3.5" = list(version_min = "3.3.0", 
                                      version_max = "99.99.99", 
                                      path = "bin"))), "devtools")
```
Then try the `install_github` command again.
- If you get an error about package `stringi` try typing `install.packages('stringi',type='win.binary')` then re-run the commands above.
  

#### Mac issues

  - you will more than likely need to install the command line developer tools. Instructions for this are available from (http://osxdaily.com/2014/02/12/install-command-line-tools-mac-os-x/). There is a larger set of tools called Xcode which is available from the Apple Store, but this is unnecessarily large and we don't recommend it unless you know you need it.
  - related to the installation of the command line tools, you may get errors such as `make: gfortran: no such file or directory mac` which relate to missing information for the calling of C++ and FORTRAN compilers. The solution is to download and [install the required GFORTRAN](https://thecoatlessprofessor.com/programming/cpp/r-compiler-tools-for-rcpp-on-macos/) libraries. The instructions here are not very clear, and I will try to record a short podcast running through them.
  - If you are on a Mac you might need to install X11 from [http://www.xquartz.org](http://www.xquartz.org). 
  - If you get an error about package `stringi` try typing `install.packages('stringi',type='mac.binary')`. Then re-run the commands above
  <!-- - If you get some errors about `pandoc` when installing simmr leave out the `build_vignettes=TRUE` argument -->


If you run into any other problems please drop us a line at <andrew.parnell@mu.ie> (for simmr/MixSIAR queries) or <a.jackson@tcd.ie> (for SIBER/SIDER queries).

