---
title: "Pre-requisites"
author: "Andrew Parnell and Andrew Jackson"
output: html_document
---

In preparation for the course can all participants please do the following, preferably in the correct order as some of the steps are dependent on the preceding one.

You will need to bring your own personal computer for the duration of the course.

### Step 1

Install the following using the corresponding links

-	R: [http://www.r-project.org](http://www.r-project.org)

-	Rstudio (optional): [https://www.rstudio.com](https://www.rstudio.com)

-	JAGS: [http://mcmc-jags.sourceforge.net](http://sourceforge.net/projects/mcmc-jags/files/)

### Step 2

Now install the following packages by going into Rstudio (or R) and typing:
```{r,eval=FALSE}
install.packages(c('devtools','rjags'))
```

### Step 3

SIAR and (optionally) simmr should be installed by entering the following code in to R

```{r,eval=FALSE}
library(devtools) 
install_github("andrewljackson/siar") 
library(siar)
install_github("andrewcparnell/simmr",build_vignettes=TRUE) 
library(simmr)
```

These will install the most up to date versions of the package. Note: if you are on a Mac you might need to install X11 from [http://www.xquartz.org](http://www.xquartz.org). You may get a pop up window be asking if you want 'additional build tool' -select 'yes')

### Step 4

MixSIAR should be installed by following the instructions here: [https://github.com/brianstock/MixSIAR](https://github.com/brianstock/MixSIAR)
 
Finally if you would like to do some pre-course reading (recommended) ther are some papers which can accessed [here](https://github.com/andrewcparnell/simms_course/tree/master/papers).

### Troubleshooting

If you run into any problems please drop us a line at <andrew.parnell@ucd.ie> or <a.jackson@tcd.ie>.

