---
title: "Pre-requisites"
author: "Andrew Parnell and Andrew Jackson"
output: html_document
---

In preparation for the course please install the following, preferably in the below suggested order.

Remember you will need to bring your own personal computer for the duration of the course.

### Step 1

Install the following using the corresponding links

-	R: [http://www.r-project.org](http://www.r-project.org)

-	Rstudio (optional but recommended): [https://www.rstudio.com](https://www.rstudio.com)

-	JAGS: http://sourceforge.net/projects/mcmc-jags/files/

### Step 2

Now install the following packages by going into Rstudio (or R) and typing:
```{r,eval=FALSE}
install.packages(c('devtools','rjags'))
```

### Step 3

SIAR and (optionally) simmr should be installed by entering the following code in to R

```{r,eval=FALSE}
library(devtools) 
install_github("andrewljackson/siar") 
library(siar)
install_github("andrewljackson/SIBER") 
library(SIBER)
install_github("andrewcparnell/simmr",build_vignettes=TRUE) 
library(simmr)
```

These will install the most up to date versions of the package. You might run into the following problems:

  - If you are on a Mac you might need to install X11 from [http://www.xquartz.org](http://www.xquartz.org). 
  - You may get a pop up window be asking if you want 'additional build tools'. Please select 'yes'.
  - If you get an error about package `stringi` try typing `install.packages('stringi',type='win.binary')` if on Windows or `install.packages('stringi',type='mac.binary')` if on Mac. Then re-run the commands above
  - If you get some errors about `pandoc` when installing simmr leave out the `build_vignettes=TRUE` argument

### Step 4

MixSIAR should be installed by following the instructions here: [https://github.com/brianstock/MixSIAR](https://github.com/brianstock/MixSIAR)
 
Finally if you would like to do some pre-course reading (recommended) there are some papers which can accessed [here](https://github.com/andrewcparnell/simms_course/tree/master/papers).

### Troubleshooting

If you run into any problems please drop us a line at <andrew.parnell@ucd.ie> or <a.jackson@tcd.ie>.

