---
title: "The statistical model behind simmr (and SIAR)"
author: Andrew Parnell \newline \texttt{andrew.parnell@mu.ie}   \newline \vspace{1cm}
  \newline \includegraphics[width=3cm]{../maynooth_uni_logo.jpg}
output:
  beamer_presentation:
    includes:
      in_header: ../header.tex
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'pdf')
par(mar=c(3,3,2,1), mgp=c(2,.7,0), tck=-.01,las=1)
```
---

## Learning outcomes:
- Understand the statistical model behind simmr/SIAR
- Know how to run a model in simmr/SIAR and check that it works
- Be able to follow the technical details of the 2010 SIAR Plos ONE paper

## Our simple SIMM

- In the last class we had a simple SIMM defined via:
$$ y_i \sim N\left(\sum_{k=1}^2 p_ks_k,\sigma^2\right)$$
with $s_k \sim N(\mu_{s_k},\sigma_{s_k}^2)$, $p_1 \sim U(0,1)$ and $\sigma \sim U(0,100)$
- Here $y_i$ is the isotope value, $s$ are the source values, $p$ are the dietary proportions, and $\sigma$ is the residual standard deviation
- The goal is to estimate the $p$ and its uncertainty. The other parameters can be considered nuisance parameters

## Expanding the simple SIMM

- This SIMM is currently too simplistic. We need to expand it by:
    - increasing the number of food sources
    - including trophic enrichment factors (TEFs)
    - including concentration dependence
    - allowing for multiple isotopes
    - allowing for richer source sampling by consumers
- If we include all of these factors we end up with the simmr/SIAR model
- We will take them in turn and add them into our JAGS code

## Reminder: the SIAR geese data

```{r,include=FALSE}
library(siar)
```
```{r}
data(geese1demo,sourcesdemo, correctionsdemo, concdepdemo)
head(geese1demo,3)
sourcesdemo
```

## Plotting the data {.smaller}
\small
A plot in isotope space:
```{r,results='hide',message=FALSE, echo = FALSE, warning=FALSE}
library(simmr)
mix = geese1demo
colnames(mix) = c('d13C','d15N')
s_names=as.character(sourcesdemo[,1])
s_means = sourcesdemo[,c(2,4)]
s_sds = sourcesdemo[,c(3,5)]
c_means = correctionsdemo[,c(2,4)]
c_sds = correctionsdemo[,c(3,5)]
conc = concdepdemo[,c(2,4)]
# Load
simmr_in = simmr_load(mixtures=mix,
                     source_names=s_names,
                     source_means=s_means,
                     source_sds=s_sds,
                     correction_means=c_means,
                     correction_sds=c_sds,
                     concentration_means = conc)
```

```{r, warning=FALSE, fig.height = 5}
plot(simmr_in)
```

## Including multiple sources

- Adding in multiple sources to the likelihood means having more terms in the sum:
$$ y_i \sim N\left(\sum_{k=1}^K p_ks_k,\sigma^2\right)$$
- In the above we have $K$ sources and hence $K$ dietary proportions
- We also now need $K$ source prior distributions
- The tricky part about adding in multiple proportions is the prior distribution

## Priors for constrained dietary proportions

- We must have $\sum_{k=1}^K p_k = 1$ so any prior distribution we place on the $p$s must satisfy this restriction
- (You will often hear values restricted in sum referred to as a _simplex_)
- Luckily there is a distribution known as the _Dirichlet_ which is suitable for restricted sum parameters
- The Dirichlet has one parameter for each proportion $\alpha_1,\ldots,\alpha_K$. The larger the $\alpha$ value the larger prior weight that dietary proportion will be given
- Setting all the $\alpha$ values to 1 is equivalent to the simplex uniform distribution, i.e. a prior assumption that all sources are consumed equally

## JAGS SIMM with a Dirichlet prior
```{r,include=FALSE}
library(siar)
library(R2jags)
data(geese1demo); data(sourcesdemo)
consumers = geese1demo[,2]
set.seed(4)
```

\small
```{r,results='hide'}
model_code ='
model {
  for(i in 1:N) { y[i] ~ dnorm(inprod(p,s),sigma^-2) }
  p ~ ddirch(alpha)
  for(k in 1:K) { s[k] ~ dnorm(s_mean[k],s_sd[k]^-2) }
  sigma ~ dunif(0,100)
}
'
sources = sourcesdemo[,4:5]
data=list(y=consumers,s_mean=sources[,1],s_sd=sources[,2],
          N=length(consumers),K=nrow(sources),
          alpha=rep(1,nrow(sources)))
set.seed(123)
model_run = jags(data = data,
                 parameters.to.save = c("p", "sigma"),
                 model.file = textConnection(model_code))
```
This is now running with all 4 sources

## Results

\small

- We can explore/plot results with `summary(output)`, `plot(output)`, and also run multiple chains, form predictive distributions, check convergence, etc
- One important thing to note is that the fitting method (MCMC) produces a joint posterior distribution of the dietary proportions. This means that each set of samples will sum to 1:
```{r}
head(model_run$BUGSoutput$sims.matrix,4)
```
- The key implication of this is that, aside from exploring the _marginal_ posterior distributions (with means, sds, etc) we can explore the _joint_ uncertainty of the dietary proportions


## A joint plot of the posterior dietary proportions

```{r,include=FALSE}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- cor(x, y)
    txt <- format(c(r, 0.123456789), digits = digits)[1]
    txt <- paste0(prefix, txt)
    if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * abs(r))
}

```

```{r, fig.height = 5}
out_2 = model_run$BUGSoutput$sims.list$p
colnames(out_2) = sourcesdemo[,1]
pairs(out_2, lower.panel = panel.smooth, 
      upper.panel = panel.cor)
```

## Trophic enrichment factors and concentration dependence {.smaller}

- Trophic enrichment factors ($c$) and concentration dependence ($q$) represent adjustments to the source values to account for various measurement effects
- We can include them by expanding the likelihood:
$$ y_i \sim N \left( \frac{\sum_{k=1}^K p_k q_k (s_k+c_k)}{\sum_{k=1}^K p_k q_k}, \sigma^2\right) $$
- The extra part on the denominator is needed so that the dietary proportions still sum to 1
- The prior for $c_k$ comes from external data and are given normal distributions like the source values
- In SIAR the concentration dependencies must be less than 1 (given as proportions) and are treated as fixed. You could use a strong Dirichlet prior on these instead

## Including TEFs and CD - JAGS model {.smaller}

\small
```{r,results='hide'}
model_code ='
model {
  for(i in 1:N) { 
    y[i] ~ dnorm(inprod(p*q,s+c)/inprod(p,q),sigma^-2) 
  }
  p ~ ddirch(alpha)
  for(k in 1:K) { 
    s[k] ~ dnorm(s_mean[k],s_sd[k]^-2) 
    c[k] ~ dnorm(c_mean[k],c_sd[k]^-2) 
  }
  sigma ~ dunif(0,100)
}
'
data(concdepdemo); data(correctionsdemo)
data=list(y=consumers,s_mean=sources[,1],s_sd=sources[,2],
          c_mean=correctionsdemo[,4],c_sd=correctionsdemo[,5],
          q=concdepdemo[,4],N=length(consumers),K=nrow(sources),
          alpha=rep(1,nrow(sources)))
model_run = jags(data = data,
                 parameters.to.save = c("p"),
                 model.file = textConnection(model_code))
```

## Notes on the TEF and CD model

- If you run this, you'll find that convergence isn't quite as neat and it starts to get a bit slower
- Although it's a nuisance parameter, saving `sigma` is often a good idea because a large value indicates a poorly fitting model (usually also seen in the iso-space plot)
- The model will also create posterior distributions for $s$ and $c$, though these are usually pretty similar to the prior, as there isn't much information about their values in the data

## Adding extra isotopes

- If we have extra isotopes we can just list the likelihood twice, once for each value of the isotope. Only the dietary proportions are 'shared' between the isotopes
- Now write $y_{ij}$ as the consumer values for observation $i$ on _isotope_ $j$, where $j=1,\ldots,J$ 
- We now have source values $s_{jk}$, TEF values $c_{jk}$, concentration dependencies $q_{jk}$, and each isotope has its own residual standard deviation $\sigma_j$
- The likelihood is now:
$$ y_{ij} \sim N \left( \frac{\sum_{k=1}^K p_k q_{jk} (s_{jk}+c_{jk})}{\sum_{k=1}^K p_k q_{jk}}, \sigma_j^2\right) $$


## Richer source sampling {.smaller}

- The model we've been fitting up to now assumes that all individuals sample the same source value $s_k$ for each source and isotope. This is unrealistic
- A better model has each individual sampling a different source value from the source prior distribution, i.e. we now have $s_{ik}$ (or $s_{ikj}$ with multiple isotopes)
- The JAGS code becomes:
```{r,eval=FALSE}
for(k in 1:K) { 
  for(i in 1:N) {
    s[i,k] ~ dnorm(s_mean[k],s_sd[k]^-2) 
  }
}
```
- We can do the same with the concentration dependence values
- In fact with a bit of clever maths we can remove (_marginalise over_) the $s_{ik}$ values to get a simpler model with fewer parameters.

## The full simmr/SIAR model

- Using the trick mentioned on the last slide, we end up with a full model which looks like this:
$$ y_{ij} \sim N \left( \frac{\sum_{k=1}^K p_k q_{jk} (\mu_{s,jk}+\mu_{c,jk})}{\sum_{k=1}^K p_k q_{jk}}, \frac{\sum_{k=1}^K p_k^2 q_{jk}^2 (\sigma_{s,jk}^2+\sigma_{c,jk}^2)}{(\sum_{k=1}^K p_k q_{jk})^2} +  \sigma_j^2\right) $$
- This model has a more complicated likelihood, but removes the extra $s$ and $c$ parameters

## Full SIAR model: JAGS code
```{r,results='hide'}
model_code ='
model {
  for (i in 1:N) { 
    for (j in 1:J) {
      y[i,j] ~ dnorm(inprod(p*q[,j], s_mean[,j]+c_mean[,j]) / 
        inprod(p,q[,j]), var_y[j]^-1)
    }
  }
  p ~ ddirch(alpha)
  for(j in 1:J) {
    var_y[j] <- inprod(pow(p*q[,j],2),s_sd[,j]^2+c_sd[,j]^2)/pow(inprod(p,q[,j]),2)
      + pow(sigma[j],2)
  }
  for(j in 1:J) { sigma[j] ~ dunif(0,100) }
}
'
```

## Full simmr/SIAR model: R code

\small
```{r,results='hide'}
sources = as.matrix(sourcesdemo[,2:5])
tefs = as.matrix(correctionsdemo[,2:5])
cd = as.matrix(concdepdemo[,c(2,4)])
data=list(y=geese1demo,s_mean=sources[,c(1,3)],
          s_sd=sources[,c(2,4)],
          c_mean=tefs[,c(1,3)],c_sd=tefs[,c(2,4)],
          q=cd,N=nrow(geese1demo),
          J=ncol(geese1demo),alpha=rep(1,nrow(sources)))
model_run = jags(data = data,
                 parameters.to.save = c("p", "sigma"),
                 model.file = textConnection(model_code),
                 DIC = FALSE)
```

## Summary of posterior dietary proportions

\small
```{r,fig.align='center'}
out_2 = model_run$BUGSoutput$sims.matrix
colnames(out_2) = c(as.character(sourcesdemo[,1]),'SD1','SD2')
t(round(apply(out_2,2,quantile,probs=c(0.025,0.5,0.975)),2))
```
Some of these proportions are quite imprecise: perhaps see better with matrix plot?

## Running SIAR/simmr

- The SIAR/simmr R packages run exactly this model with a few extra tweaks
- It contains a slightly optimised algorithm as JAGS sometimes gets a bit stuck on harder data sets. It's also much faster than JAGS for complicated problems
- It allows for direct plotting of the data in isotope space and $p$-space (i.e. dietary proportion space - `pairs` plots)
- It allows for changing the $\alpha$ values to put in proper prior information
- It includes convergence checking 
- Most of this covered in the practical this afternoon

## simmr version

- `simmr` is a much more elegantly written version of SIAR with neater plots and many more features
- Four steps to run a `simmr` model
  1. Call `simmr_load` to load in the data
  1. Call `plot` to see the iso-space plot
  1. Call `simmr_mcmc` to run the model
  1. Call `plot` or `summary` to access the output
- `simmr` has further features to combine sources and to compare dietary proportions

## simmr code

```{r, echo = FALSE}
library(simmr)
mix = geese1demo
colnames(mix) = c('d13C','d15N')
s_names=as.character(sourcesdemo[,1])
s_means = sourcesdemo[,c(2,4)]
s_sds = sourcesdemo[,c(3,5)]
c_means = correctionsdemo[,c(2,4)]
c_sds = correctionsdemo[,c(3,5)]
conc = concdepdemo[,c(2,4)]
```

```{r, eval = FALSE}
# Load
simmr_in = simmr_load(mixtures=mix,
                     source_names=s_names,
                     source_means=s_means,
                     source_sds=s_sds,
                     correction_means=c_means,
                     correction_sds=c_sds,
                     concentration_means = conc)
# Iso-space plot
plot(simmr_in)
# MCMC run
simmr_out = simmr_mcmc(simmr_in)
# Box-plots
plot(simmr_out, type = 'boxplot')
```

## Summary

- The simmr and SIAR models are just complicated versions of linear regression
- The response is multivariate and the prior distributions on some of the parameters have to be constrained to sum to 1
- It used to be the case that JAGS was slow and couldn't run SIMM-type models. This is no longer true. You can fit much richer models in JAGS (and now MixSIAR) than with SIAR/simmr
- More details on running simmr in the practical next

