---
title: "MixSIAR script example as Rnotebook"
output: html_notebook
---

Here I recreate the mixsiar_script.R example but in a Rnotebook. Additionally I have grabbed the actual data files and put them in a folder `wolves_data/` within the folder where this Rmd file sits so that the analysis more closely resembles the analysis you might make with your own data.

## Setup 

Load libraries.

```{r}
library(MixSIAR)
library(tidyr) # For pivoting later in new output_jags
library(ggplot2)
```

## Load in the data

First load the mixtures, i.e. the consumers.

```{r}
# Load into MixSIAR
mix = load_mix_data(
  filename="wolves_data/wolves_consumer.csv", 
  iso_names=c("d13C","d15N"), 
  factors=c("Region","Pack"), 
  fac_random=c(TRUE,TRUE), 
  fac_nested=c(FALSE,TRUE), 
  cont_effects=NULL)
```

load the Source data

```{r}
# Load the source data
source = load_source_data(
  filename="wolves_data/wolves_sources.csv",
  source_factors="Region", 
  conc_dep=FALSE, 
  data_type="means", 
  mix)
```

Load the trophic discrimination factors.

```{r}
# Load the discrimination/TDF data
discr = load_discr_data(
  filename="wolves_data/wolves_discrimination.csv", 
  mix)
```

## Plot the data

```{r}
# Isospace plot 
plot_data(filename="isospace_plot", 
          plot_save_pdf = FALSE, 
          plot_save_png = FALSE,
          mix = mix, 
          source = source, 
          discr = discr)

```

Plot the same data but with a colour-blind friendly palette and hide the legend. 

```{r}
# Alternative plot saving it and editing it
p <- plot_data(filename="isospace_plot", 
          plot_save_pdf = FALSE, 
          plot_save_png = FALSE,
          mix = mix, 
          source = source, 
          discr = discr, return_obj = TRUE)

# We can now augment the ggplot object with our own 
# customisations.
p + 
  theme(legend.position = "None") + 
  scale_color_viridis_d()
```

## Use MixSIAR to construct the JAGS model 

```{r}
# Write the model out
model_filename = "MixSIAR_model.txt"   # Name of the JAGS model file
write_JAGS_model(model_filename, 
                 resid_err = TRUE, 
                 process_err = TRUE, 
                 mix, source)
```

## Run a very short test run of the model 

```{r}
# Test run
jags.1 = run_model(run="test", 
                   mix, source, discr, 
                   model_filename, 
                   alpha.prior = c(1,1,1), 
                   resid_err = TRUE, 
                   process_err = TRUE)
```


## Run a longer version but still very short

N.B. This is not evaluated by default in case it takes a long time on your machine. Change chunk option `eval = TRUE` or copy and paste to the console window.

```{r, eval = FALSE}
# Long run - change run to 'normal' 
# if you have 5 hours to kill
jags.2 = run_model(run="very short", 
                   mix, source, discr, 
                   model_filename, 
                   alpha.prior = 1, 
                   resid_err = TRUE, 
                   process_err = TRUE)
```

## Explore the output

```{r}
out <- output_JAGS(jags.1, 
                   mix = mix, 
                   source = source,
                   c('summary_diagnostics'))
```

Print and plot all (most?) of the optional output. 

N.B. not run by default. Change chunk option `eval = TRUE` or copy and paste to the console window.

```{r}
out <- output_JAGS(jags.1, 
                   mix = mix, 
                   source = source,
                   c('summary_diagnostics',
                     'summary_statistics',
                     'summary_quantiles',
                     'plot_global',
                     'plot_factors'))
```






