---
title: "Source Aggregation"
author: "Andrew L Jackson"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Create some simulated data for us to work with
This is essentially the example used in Fry, B. 2013. Alternative approaches for solving under-determined isotope mixing problems. *MEPS*.

Here we use the package `simmr` to explore the two alternatives to aggregating sources in mixing models. 

```{r simulate-data}
library(simmr, quietly = TRUE)

# Set the random seed so we get the same set of 
# random numbers each time we run this example.
set.seed(1)

# specify the sources
# sources <- data.frame(sources=c("A","B","C","D"),
#                         muC=c(-5,-5,5,5),sdC=c(1,1,1,1),
#                         muN=c(-5,5,5,-5),sdN=c(1,1,1,1))

# specify the source names
S_names = c("A","B","C","D")

# specify the source means by binding two vectors 
# together into a matrix by columns.
S_means = cbind(c(-5,-5,5,5), c(-5,5,5,-5))

# specify the source standard deviations
S_sds = cbind(c(1,1,1,1), c(1,1,1,1))

# speficy the consumer data at the origin
# Ten consumers for this example around 0 with small sd of error.
consumers <- cbind(dC = rnorm(n = 10, mean = 0, sd = 0.1),
                   dN = rnorm(n = 10, mean = 0, sd = 0.1) )

# and create the simmr object
# here we have no TDFs or concentration values
simmr_in <- simmr_load(mixtures = consumers,
                       source_names = S_names,
                       source_means = S_means,
                       source_sds = S_sds)

```

Now we can plot the data to visualise our system and the output,

```{r plot-isotope-data}

plot(simmr_in)

```

## Fit the SIMM

We can fit a simmr model using the defaults, and here supress the output using the `results='hide'` option in the chunk.

```{r, results='hide'}
simmr_out = simmr_mcmc(simmr_in)
```

And we should always check for convergence.

```{r}
# a summary table of convergence diagnostics
summary(simmr_out,type='diagnostics')

# plot the posterior predictive power 
posterior_predictive(simmr_out)
```

The results of this toy example are not expected to be overly helpful or meaningful.

```{r}
summary(simmr_out,type='statistics')
summary(simmr_out,type='quantiles')
```

Plot the estimates of the dietary proportions

```{r}
# Plot the a priori aggregated diet estimatess
plot(simmr_out, type = "density")
```


Plot the covariance between the estimated dietary proportons in the posterior.

```{r plot-posterior-cov}

plot(simmr_out, type = "matrix")

```



## __*A priori*__ aggregation

We combine the sources C and D before we run the model as is sometimes suggested. We do this by taking the mean of the means, and we square the SDs to make them variances, then add them, and then square-root them to turn them back into SDs again. We dont need to change the consumer data in any way.

```{r a-priori}

# specify the source names
S_names_a_priori = c("A","B","CD")

# take the mean of the last two sources C and D
S_means_a_priori = rbind(S_means[1:2,], colMeans(S_means[3:4,]) )

# square the sds for sources C and D to convert to variance,
# sum them and convert back to sd
S_sds_a_priori = rbind( S_sds[1:2,], colSums(S_sds[3:4,]^2) ^ 0.5 )

# and create the new simmr object
# here we have no TDFs or concentration values
simmr_in_a_priori <- simmr_load(mixtures = consumers,
                       source_names = S_names_a_priori,
                       source_means = S_means_a_priori,
                       source_sds = S_sds_a_priori)




```

Plot this newly combined data. 

```{r}
# plot the raw data for the a priori aggregated example
plot(simmr_in_a_priori)
```

Run the model on the a priori combined data.

```{r}
# fit the a priori aggregated models
simmr_out_a_priori = simmr_mcmc(simmr_in_a_priori)

```
And plot

```{r}
# Plot the a priori aggregated diet estimatess
plot(simmr_out_a_priori, type = "density")
```

... and now apparently we are very sure about the contributions of all sources to the diet. There is some correlation between A and B since they need to balance each other out in combination to yield a dB value of 0. You would now incorrectly assume that  CD represents pretty much a guaranteed 43% of the diet.

```{r plot-posterior-cov-prior-agg}

plot(simmr_out_a_priori, type = "matrix")

```

## __*A posteriori*__ aggregation
Far more honest is to fit the model as before, with the sources as we believe them to be *a priori* and then simply add our prortions together from the posterior distribution. This is made very easy in SIMMR and also MixSIAR with dedicated functions. In fact, MixSIAR also allows easier a priori aggregation by hiding the routine outlined in the preceding section.

```{r a-posteriori}
# combine sources C and D which are in positions 3 and 4
simmr_out_a_posteriori <- combine_sources(
  simmr_out, 
  to_combine = simmr_out$input$source_names[c(3,4)], 
  new_source_name = "CD")

# Plot the a posteriori aggregated diet estimatess
plot(simmr_out_a_posteriori, type = "density")


```
This result fits much better with what we would predict: that if the model is still not sure about the contribution of the four sources to the mixture, but that it is pretty sure that on average, 50% of the diet is comprised of both C and D. This concept continues until the model is entirely certain, with no error, that the diet is wholly 100% of A+B+C+D.

One thing to experiment with here is the use of the Jeffrey's prior of `c(0.25, 0.25, 0.25, 0.25)` in place of the default vague prior `c(1, 1, 1, 1)`. This is the nub of the criticism levelled at the SIMMs by Brett, M. 2016. Resource polygon geometry predicts Bayesian stable isotope mixing model bias. MEPS.


