---
title: "Source Aggregation"
author: "Andrew L Jackson"
date: "27 February 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Create some simulated data for us to work with
This is essentially the example used in Fry, B. 2013. Alternative approaches for solving underdetermined isotope mixing problems. MEPS.

```{r simulate-data}
library(siar, quietly = TRUE)

set.seed(1)

# specify the sources
sources <- data.frame(sources=c("A","B","C","D"),
                        muC=c(-5,-5,5,5),sdC=c(1,1,1,1),
                        muN=c(-5,5,5,-5),sdN=c(1,1,1,1))

# speficy the consumer data at the origin
# Ten consumers for this example around 0 with small sd of error.
consumers <- data.frame(C = rnorm(10, 0, 0.1),
                        N = rnorm(10, 0, 0.1) )

```

Now we need to run the model and we can then plot the data to visualise our system and the output,

```{r run-model}
model1 <- siarmcmcdirichletv4(consumers, sources)

siarplotdata(model1,iso=c(1,2))

siardensityplot(model1$output[, 1:4], xticklabels=c('A','B','C','D'),
                main = "All 4 sources")

```

## __*A priori*__ aggregation
We combine the sources C and D before we run the model as is sometimes suggested. We do this by taking the mean of the means, and we square the SDs to make them variances, then add them, and then square-root them to turn them back into SDs again.

```{r a-priori}
combined.sources <- data.frame(sources=c("A","B","CD"),
                        muC = c(sources$muC[c(1,2)], mean(sources$muC[c(3,4)])),
                        sdC = c(sources$sdC[c(1,2)], sqrt(sum(sources$sdC[c(3,4)]^2))),
                        muN = c(sources$muN[c(1,2)], mean(sources$muN[c(3,4)])),
                        sdN = c(sources$sdN[c(1,2)], sqrt(sum(sources$sdN[c(3,4)]^2)))
                        )

model2 <- siarmcmcdirichletv4(consumers, combined.sources)

#dev.new()
siarplotdata(model2,iso=c(1,2))


#dev.new()
#siarhistograms(model2)

#dev.new()
siardensityplot(model2$output[ , 1:3],
                xticklabels=c('A','B','C+D'),
                main = "Sources C and D combined a priori")


siarhdrs(model2)

siarmatrixplot(model2)

```

... and now apparently we are very sure about the contributions of all sources to the diet. There is some correlation between A and B since they need to balance each other out in combination to yield a dB value of 0. You would now incorrectly assume that  CD represents pretty much a guaranteed 43% of the diet.

## __*A posteriori*__ aggregation
Far more honest is to fit the model as before, with the sources as we believe them to be *a priori* and then simply add our prortions together from the posterior distribution.

```{r a-posteriori}
CD.posteriori.aggregated <-  cbind(model1$output[,c("A","B")],
                                   rowSums(model1$output[,c("C","D")]))

BCD.posteriori.aggregated <-  cbind(model1$output[,c("A")],
                                   rowSums(model1$output[,c("B","C","D")]))


# this figure illustrates how uncertain we are in the estiamtes, and 
# it differs starkly with the relatively much more certain results stemming 
# from the a priori aggregation.
#dev.new()
siardensityplot(CD.posteriori.aggregated,
                xticklabels=c('A','B','C+D'),
                main = "Sources C and D combined a posteriori")

pairs(CD.posteriori.aggregated,
      diag.panel = panelhist,
      lower.panel = panelcor,
      upper.panel = panelcontour)
```
This result fits much better with what we would predict: that if the model is still not sure about the contribution of the four sources to the mixture, but that it is pretty sure that on average, 50% of the diet is comprised of both C and D. This concept continues until the model is entirely certain, with no error, that the diet is wholly 100% of A+B+C+D.

One thing to experiment with here is the use of the Jeffrey's prior of `c(0.25, 0.25, 0.25, 0.25)` in place of the default vague prior `c(1, 1, 1, 1)`. This is the nub of the criticism levelled at the SIMMs by Brett, M. 2016. Resource polygon geometry predicts Bayesian stable isotope mixing model bias. MEPS.


